An Efficient Forecasting Approach for the Reduction of Boundary Effects in Time-Frequency Representations

### 2020-07-02 Notes/BoundEffRed.tex
paper and code

\title{An Approach for the Reduction of Boundary Effects in Time-Frequency Representations}

@@

\begin{abstract}
Time-frequency representations are intrinsically subject to the boundary effects. As a consequence, the structures of signals that are highlighted by the representations are garbled when approaching the boundaries of the time-frequency domain. In this paper, we propose an approach for the reduction of such boundary effects, whose objective is to be able to be used in real time. That one relies on an extension of the analyzed signal obtained by a forecasting technique. In the case of the study of a class of locally oscillating signals, we provide a theoretical study of the performance of our approach. Following a numerical verification of the performance of the algorithm, we implement it on real-life biomedical signals.
\end{abstract}

@@

\section{Introduction}
\label{se:introduction}
\IEEEPARstart{I}{n} any digital acquisition system, the study and the interpretation of the measured signals generally require the use an analysis tool, which enable to point out the useful characteristics of the signal. This problem arises for the analysis of signals as varied as audio~\cite{Stowell18computational,Muller11signal}, mechanical~\cite{Peng02vibration} or biomedical signals~\cite{Akay96detection}. For instance, any  measured biomedical signal, such as photoplethysmogram (PPG), may not be interpreted as it is, from its run chart. An analysis tool would make possible the extraction of some useful characteristics such as heart rate or blood pressure.

In general, the observed quantities are produced by transient phenomena, that can vary rapidly and irregularly. As a consequence, the measured signals exhibit nonstationary behavior. In order to adapt the analysis to nonstationarities, local spectral analysis is generally performed~\cite{Stoica05spectral,Matz97generalized}. The short time Fourier transform~\cite{Grochenig01foundations} (STFT) is a typical tool which is build that way, and enables the determination of the local frequency content of a nonstationary signal.

Windowing is a common method for performing local analysis. Among others, STFT, wavelet transform~\cite{Mallat99wavelet}, synchrosqueezing transform (SST), and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is is localized around the origin, and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ rely on the evaluation of the following dot product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ near the boundaries of the interval $I$. Indeed, at these points, half of the information is missing. Consequently the results of the dot product~\ref{eq:windowing} is disturbed by this lack of information. This phenomenon is the so-called \emph{boundary effect}. We display, on the left of Fig.~\ref{fig:ex.intro} the result of the SST of a PPG (see section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on both sides of this representation. Indeed, while in the major central part of the image, clear lines stand out, they become blurred as they approach the left and right boundaries of the image. A zoom on the right boundary of this SST, displayed on the right of Fig.~\ref{fig:ex.intro}, emphasizes the result of boundary effects. More generally, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from its time-frequency representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries.

\begin{figure}
\centering
%\includegraphics[width=.48\textwidth]{PPGsig.eps}
\includegraphics[width=.48\textwidth]{SSTintro.eps}
\includegraphics[width=.48\textwidth]{zoomSSTintro.eps}
\caption{Synchrosqueezing transform of a PPG (top) with a zoom on its right boundary (bottom). }
\label{fig:ex.intro}
\end{figure}

The purpose of this paper is to provide a fast algorithm allowing to tackle boundary effects and limit the disruptions it may cause on the signals representations based on windowing. Attempts to minimize the boundary effects generally consists in softening the discontinuity on signals edges. For instance, judicious choices of analysis windows whose support does not interfere with the boundary  points can minimize the occurrence of aberrant patterns near the boundaries of the time-frequency plane~\cite{Chui92wavelets,Depczynski99fast}. These techniques therefore do not make it possible to reduce the boundary effects of any time-frequency representations, and for any analysis window. For this purpose, a natural idea consists in carrying out a preliminary step to the analysis allowing to extend the signal beyond its boundaries. Hence, to implement our boundary effects reduction procedure, we proceed in two steps.
\begin{enumerate}
\item \emph{Extend the signal by forecasting it.} The aim is to use a dynamic model to predict the values taken by the measured signal outside the measurement interval. Then, once this operation is done, we have access to an extended signal defined on a larger interval $I_\Delta$, where $\Delta$ denotes the size of the extension on both boundaries of $I$.
\item \emph{Run the local analysis tool on the extended signal.} Assuming that the support of the analysis window is smaller that $2\Delta$, the local analysis near the boundary of $I$ is now possible without lack of information thanks to knowledge brought by the extension on both sides. 
\end{enumerate}
Thus, assuming that the quality of the extension step is sufficient, the representation obtained that way will be less sensitive to the boundary effects than the result of the analysis tool applied directly to the non-extended signal.

Various extension schemes have been proposed. On one hand, there exists simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. On the other hand, there exists extension schemes based on physically relevant dynamical models, such as the Extented Dynamic Mode Decomposition~\cite{Williams15data} (EDMD), the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR), the TBATS algorithm~\cite{DeLivera11forecasting}.  While the second class of extension schemes give better extended signals than the first class, they generally have a great computational cost. We propose here a fast extension algorithm based on a simple dynamical model, in order to optimize the trade-off between the extension quality and the computational cost. Furthermore, we prove that the dynamical model we consider is sufficient to forecast locally oscillating signals.

Therefore, in section~\ref{se:algo}, we provide an extension method based on a linear dynamic model. We derive the corresponding algorithm for boundary effects reduction. In section~\ref{se:theoretical}, we show that, even though we consider a simplistic dynamic model, it is sufficient to extend signals takings the form of sums of sine waves. An evaluation of the theoretical performance of our algorithm on a class of signals, the sums of sine waves, is given in section~\ref{se:theoretical}. In section~\ref{se:results}, we compare our extension method with more sophisticated methods such as EDMD, GPR, or TBATS. We show that our algorithm gives fast results of reasonable quality. Finally, we evaluate the performance of our boundary effects reduction algorithm on biomedical signals such as respiratory signals, and compare it to the theoretical results.

@@

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the boundary effects reduction on time-frequency representations. This method is based on an extension of the signal obtained by forecasting. We have shown theoretically that the chosen dynamic model is sufficient for the extension of signals formed by a sum of sine waves whose instantaneous frequencies and amplitudes vary slowly. Moreover, the numerical results show that this strategy allows us to switch to a real-time implementation of {\sf BoundEffRed} is possible, due to the low computation time, unlike other pre-existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many time-frequency representations. The MATLAB code and datasets used to produce the numerical results of this paper are available online at the following address:
\begin{center}
    {\small\url{https://github.com/AdMeynard/BoundaryEffectsReduction}}
\end{center}
Other numerical examples are also included, such as the application of {\sf BoundEffRed} to other types of biomedical signals such as ECGs.

Various improvements can be considered to make the algorithm more robust to certain situations. In particular, we have noticed that when the observed signal does not oscillate, the forecasting is no longer relevant, and outlier time-frequency representations are obtained. A preliminary step should then be added to the algorithm to select the portions of the signal that exhibit oscillations, in order to reject the portions for which accurate time-frequency analysis is practically useless. A more fundamental improvement would be to perform the prediction step directly in the time-frequency plane, with the aim of specifically predicting the evolution of each instantaneous frequency. This would lead to seeing a time-frequency representation as a set of interdependent time series (corresponding to each frequency channel).

>>>>

\title{An Efficient Forecasting Approach for the Reduction of Boundary Effects in Time-Frequency Representations}

@@

\begin{abstract}
Time-frequency (TF) representations of time series are intrinsically subject to the boundary effects. As a consequence, the structures of signals that are highlighted by the representations are garbled when approaching the boundaries of the TF domain. In this paper, for the purpose of real-time TF information acquisition for nonstationary oscillatory time series, we propose a numerically efficient approach for the reduction of such boundary effects. The solution relies on an extension of the analyzed signal obtained by a forecasting technique. In the case of the study of a class of locally oscillating signals, we provide a theoretical guarantee of the performance of our approach. Following a numerical verification of the performance of the algorithm, we validated its performance on various real-life biomedical signals.
\end{abstract}

@@

\section{Introduction}
\label{se:introduction}
\IEEEPARstart{I}{n} any digital acquisition system, the study and the interpretation of the measured signals generally require an analysis tool, which enables researchers to point out the useful characteristics of the signal. The need for signal analysis arises from various signals, ranging from audio~\cite{Stowell18computational,Muller11signal}, mechanical~\cite{Peng02vibration} or biomedical signals~\cite{Akay96detection}. In this paper, we focus on oscillatory time series. For instance, biomedical signal, such as photoplethysmogram (PPG), contain several characteristics, such as respiratory rate or blood pressure, that cannot be interpreted from its running chart in the time domain. An analysis tool would make possible the extraction of these useful characteristics.

Usually, the measured signals exhibit nonstationary behavior, and the observed quantities might be interfered by transient phenomena that can vary rapidly and irregularly. For example, the signal might oscillate fast with large amplitude at one moment, and then oscillate slow with small amplitude at the next moment. In order to adapt the analysis to nonstationarities, local spectral analysis is generally performed~\cite{Stoica05spectral,Matz97generalized}. The short time Fourier transform~\cite{Grochenig01foundations} (STFT) is a typical tool which is build for this purpose, and enables the determination of the local frequency content of a nonstationary signal.

Windowing is a common method for performing local analysis. Among many others, STFT~\cite{Flandrin:1999}, continuous wavelet transform (CWT)~\cite{Da1992}, synchrosqueezing transform (SST) \cite{Daubechies11synchrosqueezed}, and reassignment~\cite{Auger13time} (RS) are representations that fall back on the use of an analysis window. Let $x:I\to\RR$ denote the observed signal, where $I$ denotes the finite interval where the signal is measured. Let $g_s:\RR\to\RR$ denote the analysis window, where $s$ is a shape parameter. The support of $g_s$ is is localized around the origin, and is small with respect to $|I|$. The translation operator is $T_\tau$ defined as:
\[
T_\tau f = f(t-\tau)\ ,\quad \forall\ f:\RR\to\RR\ .
\]
Then, the local analysis of $x$ around the instant $\tau\in I$ rely on the evaluation of the following inner product:
\begin{equation}
V_x(s,\tau) = \langle x, T_\tau g_s \rangle_I \ .
\label{eq:windowing}
\end{equation}
A major shortcoming of this technique occurs when analyzing the signal $x$ {\em near the boundaries} of the interval $I$. Clearly, at these points, half of the information is missing. Consequently the results of the inner product~\eqref{eq:windowing} is distorted. This phenomenon is usually understood as the \emph{boundary effect}. We display on Fig.~\ref{fig:ex.intro} the result of the SST of a PPG (see section~\ref{ssse:ppg} for a comprehensive description). The distortion resulting from boundary effects is clearly visible on both sides of this representation. Indeed, while in the major central part of the image, clear lines stand out, they become blurred as they approach the left and right boundaries of the image. The right boundary is zoomed in and displayed on the right of Fig.~\ref{fig:ex.intro}, which emphasizes the result of boundary effects. Clearly, estimations of signal characteristics, like instantaneous frequencies~\cite{Delprat92asymptotic} or amplitudes, from this TF representation appear to be imprecisely determinable (or even likely to fail) in the vicinity of the boundaries. Moreover, this boundary effect would unavoidably limit applying these window based TF analysis tools for the real-time analysis purpose. It is thus desirable to have a solution to eliminate this boundary effect.

\begin{figure}
\centering
\includegraphics[width=.48\textwidth]{PPGsig.eps}
\includegraphics[width=.24\textwidth]{zoomSSTintro.eps}
\includegraphics[width=.24\textwidth]{zoomSSTBoundEffRed.eps}
\caption{A segment of PPG signal (top) and the right boundary of a TF representation determined by the SST without extension (bottom left), and the right boundary of a TF representation determined by the SST with the proposed boundary effect reduction algorithm by forecasting (bottom right). The window length for the SST is 12~sec.}
\label{fig:ex.intro}
\end{figure}

%The purpose of this paper is providing a fast algorithm allowing to tackle boundary effects and limit the disruptions it may cause on the signals representations based on windowing. 
Attempts to minimize the boundary effects generally consists in softening the discontinuity on signals edges. For instance, judicious choices of analysis windows whose support does not interfere with the boundary points can minimize the occurrence of aberrant patterns near the boundaries of the TF plane~\cite{Chui92wavelets,Depczynski99fast}. Due to the specific relationship of chosen analysis windows and TF analysis tool, these techniques do not make it possible to reduce the boundary effects of any TF representations, and for any analysis window. Another natural idea consists of carrying out a preliminary step of extending the signal beyond its boundaries,
%
and due to its flexibility, various extension schemes have been proposed. For example, there exists simple extension schemes that do not take into account the dynamical behavior of the signal, such as zero-padding, periodic extension, symmetric extension~\cite{Kharitonenko02wavelet,Chen95symmetric}, or polynomial extrapolation~\cite{Williams97discrete}. There also exist extension schemes based on physically relevant dynamical models, such as the Extended Dynamic Mode Decomposition~\cite{Williams15data} (EDMD), the Gaussian process regression~\cite{Rasmussen06gaussian,Roberts13Gaussian} (GPR), the Trigonometric, Box-Cox transformation, ARMA errors, Trend and Seasonal components (TBATS) algorithm~\cite{DeLivera11forecasting}, the dynamic linear models \cite{west2006bayesian}, and long short-term memory model \cite{vlachas2018data}. It is a far from exhaustive list, and we refer readers with interest to \cite{hyndman2018forecasting} for a friendly monograph on the general forecasting topic. While the second class of extension schemes give better extended signals than the first class, they generally have a great computational cost. 
%

In this paper, we propose a fast extension algorithm based on a simple dynamical model, in order to optimize the trade-off between the extension quality and the computational cost. The proposed algorithm is composed of two steps.
\begin{enumerate}
\item \emph{Extend the signal by forecasting it.} The aim is to use a dynamic model to predict the values taken by the measured signal outside the measurement interval. Then, once this operation is done, we have access to an extended signal defined on a larger interval $I_\Delta$, where $\Delta$ denotes the size of the extension on both boundaries of $I$.
\item \emph{Run the local analysis tool on the extended signal.} Assuming that the support of the analysis window is smaller that $2\Delta$, the local analysis near the boundary of $I$ is now possible without lack of information thanks to knowledge brought by the extension on both sides. 
\end{enumerate}
Thus, assuming that the quality of the extension step is sufficient, the analysis results obtained that way will be less sensitive to the boundary effects than the result of the analysis tool applied directly to the non-extended signal. 
%
We claim and prove that forecasting oscillatory signals based on the simple dynamical model combined with the simple least square approach is sufficient for reducing the boundary effects for the TF analysis, or other kernel based analysis. See Figure \ref{fig:ex.intro} for a snapshot of the result. The main benefit of this simple approach is a numerically efficient solution with theoretical guarantee for the real-time analysis purpose.

The paper is organized in the following. In section~\ref{se:algo}, we provide an extension method based on a linear dynamic model. We derive the corresponding algorithm for boundary effects reduction. In section~\ref{se:theoretical}, we show that, even though we consider a simplistic dynamic model, it is sufficient to extend signals takings the form of sums of sine waves. An evaluation of the theoretical performance of our algorithm on a class of signals, the sums of sine waves, is given in section~\ref{se:theoretical}. In section~\ref{se:results}, we compare our extension method with more sophisticated methods such as EDMD, GPR, or TBATS. We show that our algorithm gives fast results of reasonable quality. Finally, we evaluate the performance of our boundary effects reduction algorithm on biomedical signals such as respiratory signals, and compare it to the theoretical results.

@@

\section{Conclusion}
\label{se:conclusion}
In this paper, we propose an algorithm, named {\sf BoundEffRed}, for the boundary effects reduction on TF representations to achieve a real-time signal analysis. This method is based on an extension of the signal obtained by a simple-minded and numerically efficient forecasting. We have shown theoretically that the chosen dynamic model is sufficient for the extension of signals formed by a sum of sine waves. Moreover, the numerical results show that this strategy allows us to switch to a real-time implementation of {\sf BoundEffRed} is possible, due to the low computation time, unlike other existing forecasting methods. The numerical results also confirmed the robustness to noise of {\sf BoundEffRed}, as well as its ability to be applied to many time-frequency representations. 
%Other numerical examples are also included, such as the application of {\sf BoundEffRed} to other types of biomedical signals such as ECGs.

Various improvements can be considered to make the algorithm more robust to certain situations. In particular, we have noticed that when the observed signal does not oscillate {\color{red}[HT: we need to confirm this statement]}, the forecasting is no longer relevant, and the TF representations are not ideal. A preliminary step should then be added to the algorithm to select the portions without sufficient quality, in order to reject the portions for which a better forecasting scheme is needed. A more fundamental improvement would be to perform the prediction step directly in the TF plane, or even with the aim of specifically predicting the evolution of each instantaneous frequency. This would lead to seeing a TF representation as a multivariate time series (corresponding to each frequency band). We will explore these possibilities in our future work.
